White Dwarf Mergers on Adaptive Meshes\\ II. Collisions

### 2016-07-18 papers/wdmerger_II/paper.tex
Modelled --> modeled, I guess
Add 3D gravitational wave signal

Regardless of whether the suppressed burning mode works for this
particular problem, it is not physical, so in general we have
taken a different approach. If we insist that we cannot
cannot directly control the energy injection timescale, we 
must find a way to alter the sound-crossing timescale. 
We can achieve this by adding levels of refinement in 
regions that do not satisfy \autoref{eq:burning_limiter_2},
which effectively lowers the $\Delta x$ and thus the
sound-crossing time. We keep tagging zones for refinement
based on this criterion until the criterion is satisfied
on the finest level. Since the concern is regions that 
may detonate, we also tag nearby zones in a buffer region
which do not themselves satisfy the criterion,
so that a detonation in a single timestep cannot 
escape into non-refined regions. The width of the buffer 
region should thus be at least as large as the number of 
timesteps before a regridding procedure is performed.
We choose a value of two for both the number of zones in the 
buffer region and the number of steps in between regrids,
for all simulations in this paper.

While this approach may add several AMR levels 
to a simulation, we agree with \citet{kushnir:2013} that 
solving this numerical instability is crucial to avoiding
early, unphysical detonations, and so it is unavoidable
if a correct evaluation of the burning phase is desired.
A simulation that does not solve this problem in this way
(or some analogous manner) will not obtain the correct amount 
of burning, and will not converge properly with resolution. 
Fortunately, the regions where this criterion is not satisfied 
are fairly localized on the grid, so the amount of additional 
computational work is not insurmountable for a single simulation.
However, this level of accuracy may make a parameter study of
the problem challenging given current computational resources.
When resources are limited, one can still use this method and set a
limit on the number of levels of refinement possible in the simulation,
so that the refinement will still go in the right places but may not
refine all of the way to the regime of guaranteed stability.

@@

In contrast, the suppressed burning mode (modelled after \citet{kushnir:2013}) produces only about
$5 \times 10^{-4}\ \msolar$ of $^{56}$Ni. In a suppressed burn, we ensure that at every evaluation
of the right-hand-side vector for the nuclear network integration, all quantities are scaled by the
same factor, with this factor ensuring that the energy release is not large enough to permit a
spurious detonation. This seems to be equivalent to the method used in \citet{kushnir:2013}. We
reproduce their claim that this suppresses the detonation until later in the collision, after more
material has approached the stalled shock. We also observe that, for a given resolution, the simultaneous
detonations occur slightly outside the initial contact point, near the edges of the stalled shock. They
establish as the main source of error in simulations that produce too little nickel. However, we find
that the resulting detonations are not strong enough to convert a significant amount of material to NSE
conditions; instead, only QSE conditions are reached near the center of the collision, leaving a significant
amount of silicon and sulfur on the domain that is not further processed.

@@

Paper I established our methodology for calculating the gravitational wave signature
of a white dwarf merger using the post-Newtonian framework of \citet{blanchet:1990}.
While a typical merger event has a characteristic profile including a chirp and ringdown,
white dwarf collisions are noticeably distinct in their gravitational wave signatures, and
hence may be interesting targets for observation for future low frequency gravitational
wave detectors \citep{eLISA}. The strain increases monotonically for a collision until impact,
instead of oscillating back and forth at a characteristic frequency as for mergers.
The main caveat for these events compared to events
such as neutron star mergers is that due to the event being inherently less energetic,
the typical distance scale on which we could observe a white dwarf merger or collision
is the size of our own galaxy, with extragalactic events being unlikely to be
detected, at least with the first generation of space-based gravitational wave
detectors. \autoref{fig:gw_signal} is a plot of the gravitational wave strain for
the representative case of \autoref{sec:parameters}, where we disabled the normal stopping
criterion and let the simulatin run until $t = 20$. We use the convention that the
absolute scale of the strain is normalized to a distance of 10 kpc (the strain scales
inversely with the source distance). The shape of the profile agrees qualitatively
with the gravitational wave signatures predicted by \cite{loren-aguilar:2009:collisions}
and \cite{garcia-senz:2013}.

\begin{figure}
  \centering
  \includegraphics[scale=0.425]{plots/gw_signal}
  \caption{Gravitational wave strain $h_+$ for the head-on collision of two $0.64\ \msolar$
           white dwarfs. The strain is normalized to a source distance of 10 kpc. We do not
           show the other polarization, $h_\times$, as for an equal-mass head-on collision it vanishes
           identically.
           \label{fig:gw_signal}}
\end{figure}

%TODO: use a 3D collision for this?

>>>>

Regardless of whether the suppressed burning mode works for this
particular problem, it is not physical, so we include for
consideration a different approach. If we insist that we cannot
cannot directly control the energy injection timescale, we 
must find a way to alter the sound-crossing timescale. 
We can achieve this by adding levels of refinement in 
regions that do not satisfy \autoref{eq:burning_limiter_2},
which effectively lowers the $\Delta x$ and thus the
sound-crossing time. We keep tagging zones for refinement
based on this criterion until the criterion is satisfied
on the finest level. Since the concern is regions that 
may detonate, we also tag nearby zones in a buffer region
which do not themselves satisfy the criterion,
so that a detonation in a single timestep cannot 
escape into non-refined regions. The width of the buffer 
region should thus be at least as large as the number of 
timesteps before a regridding procedure is performed.
We choose a value of two for both the number of zones in the 
buffer region and the number of steps in between regrids,
for all simulations in this paper.

We agree with \citet{kushnir:2013} that solving this numerical
instability is crucial to avoiding unphysical detonations.
A simulation that does not solve this problem will not obtain
the correct amount and will not converge properly with resolution.
This may justify the addition of many AMR levels to the domain
if a correct evaluation of the burning phase is desired.
Even if one cannot afford the full resolution required by this
AMR criterion, and chooses to limit the number of levels to some
predetermined maximum based on a constraint of computing time,
the added resolution will still go to the most-needed places.
However, a significant drawback to this approach is that it
does not turn on until after an unstable burning region has
been generated, so this AMR criterion cannot help in the case
where a spurious detonation begins in a single timestep on the coarse
grid, though it will kick in immediately after that timestep
to resolve the regions where the detonation has occurred.
A possible remedy to be explored in future work is to add
refinement pre-emptively in regions where the ratio $t_s / t_e$
approaches $f_{s}$ but does not yet exceed it. Another
possibility for this case would be to track how many levels
would have been needed to prevent the numerical detonation,
based on the stability criterion, and reject the timestep and
retry it from the starting state using that many levels of refinement.
The ability to reject a timestep and retry it is functionality that
we have recently added to \castro, so we may try this in future work.

@@

In contrast, the suppressed burning mode (modeled after \citet{kushnir:2013}) produces only about
$5 \times 10^{-4}\ \msolar$ of $^{56}$Ni. In a suppressed burn, we ensure that at every evaluation
of the right-hand-side vector for the nuclear network integration, all quantities are scaled by the
same factor, with this factor ensuring that the energy release is not large enough to permit a
spurious detonation. This seems to be equivalent to the method used in \citet{kushnir:2013}. We
reproduce their claim that this suppresses the detonation until later in the collision, after more
material has approached the stalled shock. We also observe that, for a given resolution, the simultaneous
detonations occur slightly outside the initial contact point, near the edges of the stalled shock. They
establish as the main source of error in simulations that produce too little nickel. However, we find
that the resulting detonations are not strong enough to convert a significant amount of material to NSE
conditions; instead, only QSE conditions are reached near the center of the collision, leaving a significant
amount of silicon and sulfur on the domain that is not further processed.

@@

Paper I established our methodology for calculating the gravitational wave signature
of a white dwarf merger using the post-Newtonian framework of \citet{blanchet:1990}.
While a typical merger event has a characteristic profile including a chirp and ringdown,
white dwarf collisions are noticeably distinct in their gravitational wave signatures, and
hence may be interesting targets for observation for future low frequency gravitational
wave detectors \citep{eLISA}. The strain increases monotonically for a collision until impact,
instead of oscillating back and forth at a characteristic frequency as for mergers.
The main caveat for these events compared to events
such as neutron star mergers is that due to the event being inherently less energetic,
the typical distance scale on which we could observe a white dwarf merger or collision
is the size of our own galaxy, with extragalactic events being unlikely to be
detected, at least with the first generation of space-based gravitational wave
detectors.

\begin{figure}
  \centering
  \includegraphics[scale=0.425]{plots/gw_signal_2D}
  \caption{Gravitational wave strains for the head-on collision of two $0.64\ \msolar$
           white dwarfs. The strain is normalized to a source distance of 10 kpc.
           \label{fig:gw_signal_2D}}
\end{figure}

\autoref{fig:gw_signal_2D} is a plot of the gravitational wave strain for
the representative case of \autoref{sec:parameters} ($M_P = 0.64\ \msolar$, $M_S = 0.64\ \msolar$, $b = 0$),
where we disabled the normal stopping criterion and let the simulation run until $t = 20$.
We use the convention that the absolute scale of the strain is normalized to a distance
of 10 kpc (the strain scales inversely with the source distance). We show the $+$ and $\times$
polarizations for observers situated along the $x$, $y$, and $z$ axes (for the 2D case,
this corresponds to the $z$, $R$, and $\phi$ axes, respectively, in the simulation).
The shape of the profile agrees qualitatively with the gravitational wave signatures
predicted by \cite{loren-aguilar:2009:collisions} and \cite{garcia-senz:2013}.
\autoref{fig:gw_signal_3D} shows the strain prediction for a 3D collision event
with $M_P = 0.8\ \msolar$, $M_S = 0.6\ \msolar$, and $b = 0.8$. This mass combination
was chosen because \cite{loren-aguilar:2009:collisions} used the same combination for
gravitational wave signal prediction, but we cannot replicate the large impact parameter
they used on our simulation grid, so the detonation looks quite different than in their case.

\begin{figure}
  \centering
  \includegraphics[scale=0.425]{plots/gw_signal_3D}
  \caption{Gravitational wave strains for the off-center collision ($b = 0.8$) of
           $0.8\ \msolar$ and $0.6\ \msolar$ WDs. The strain is normalized to a
           source distance of 10 kpc.
           \label{fig:gw_signal_3D}}
\end{figure}
